---
title: "Survival Analysis"
author: "Monica Alexander"
date: "January 27 2021"
output: 
  pdf_document:
    number_sections: true
    toc: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, message = FALSE, warning = FALSE)
```

# Introduction

This lab will take you through some main techniques for use in survival analysis. 

## What to hand in 

Please push your Rmd and compiled document **in PDF form** to GitHub. **The questions for this week are dispersed throughout the lab.**

## Data


In this lab we're going to use the `fert` dataset in the `eha` package. This data relates to times between births for women in Sweden in the 19th century. 

As in the lecture, we're just going to look at women of parity 1: this is just demography-speak for women who have had one child already. 

So the focus of our survival analysis is the time to second birth. The variable of interest is `next.ivl`, which is the number of years until the next birth. Also of interest is the `event` variable, which tells us whether the birth happened (or whether the woman is censored). 

```{r}
library(tidyverse) # the old fave
library(survival) # useful stuff for survival analysis
library(eha) # has the dataset

data(fert)
f12 <- fert %>% as_tibble() %>% filter(parity ==1)
head(f12) 
```

Let's make a new age group variable, splitting the women by whether or not they are less than 30 years old. 

```{r}
f12 <- f12 %>% 
  mutate(age_group = ifelse(age<30, "<30", "30+"))
```


# Descriptives

## Question 1

With plots or tables, give me three observations about the times to second births. At least one of these observations should be related to differences by `age_group`.

## Answer
We have selected the largest next.ivl in age group < 30 (id 562), the largest next.ivl in age group 30+ (id 1712) and the first sample among all samples. We could find just by those three samples. The time to second births in <30 age group is larger than 30+ age group.
From the boxplot we could confirm the same conclusion, <30 age group is larger than 30+ age group.

```{r}
print(f12[f12$id %in% c(562,1712,1),])

p <- ggplot(f12, aes(x=age_group, y=next.ivl)) + 
  geom_boxplot()
p
```



# Kaplan Meier

First we will calculate the non-parametric version of the survival function. 

## `Surv` objects

`Surv` objects are set of ordered times with the censors indicated with a plus:

```{r}
survobject <- Surv(time = f12$next.ivl, event = f12$event)
head(survobject)
```


These can feed into the `survfit` function from the `survival` package to estimate the KM curve:

```{r}
fit <- survfit(Surv(next.ivl, event) ~ 1, data = f12)

fit_df <- tibble(time = fit$time, surv = fit$surv)

ggplot(aes(time, surv), data = fit_df) + 
  geom_line() + 
  ggtitle("Proportion of women who \nhave not had their second birth by time (years)") + 
  xlab("years") + ylab("S(t)")+
  theme_bw(base_size = 14)

```

## KM by hand
We can calculate Kaplan-Meier by hand fairly easily by setting up our `tibble` in the right way and calculating some new variables.

## Question 2

Fill in the gaps below (denoted by XX)

```{r}
n <- nrow(f12)

f12 <- f12 %>% 
  arrange(next.ivl) %>% # need to sort by survival times
  mutate(cumulative_people_lost = cumsum(event),
         exposure = lag(n - cumulative_people_lost, default = n),
         prob_birth = cumulative_people_lost/(cumulative_people_lost + exposure),
         prob_surv = exposure/(cumulative_people_lost + exposure),
         surv = cumprod(prob_surv))

```

If your code worked, the survival curve should be identical to what we got using `survfit`: 

(NOTE: you will need to delete the `eval=F` from this chunk and the above chunk before you compile)

```{r}
ggplot(aes(time, surv), data = fit_df) + 
  geom_line(color = "red") +
  geom_line(aes(next.ivl, prob_surv), data = f12, lty = 2)
```


## Question 3

When have 75% of the women had their second child?

## Answer:
It was 1859. Or 2.74 for the next.ivl, which could be verified on the graph.
```{r}
f12_greater_75 = f12[f12$prob_birth>0.75,]
print(f12_greater_75[1,])
```


# Piecewise Constant Hazards

Let's now estimate a PCH model, using the same cut-points as in the lecture.

## `survSplit`

To do this, we first need to get our data in the form of tracking deaths/censors in each interval. We could do this by hand, but easier with the `survSplit` function. After doing the survSplit, we then create an interval factor (for use in regression) and an interval length variable. Make sure you understand the form of this new `f12_split` and what all these new variables are. 

```{r}
cutpoints <- c(10/12, 1.25, 1.75, 2.25, seq(3,5), seq(6, 12, by = 3))
C <- length(cutpoints) + 1

f12_split <- survSplit(formula = Surv(time = next.ivl, event = event) ~ .,
                           data = f12, cut = cutpoints) %>%
  as_tibble() %>% 
  mutate(interval = factor(tstart),
         interval_length = next.ivl - tstart)
f12_split
```

Now run the regression

```{r}
fit_ind <- glm(event ~ offset(log(interval_length))-1 + interval, data=f12_split, family = "poisson")
summary(fit_ind)
```

Alternatively, we could run the Poisson regression using the sums over each interval. The results are exactly the same:

```{r}
E_k <- f12_split %>% group_by(interval) %>% summarise(E = sum(next.ivl-tstart)) %>% select(E) %>% pull()
D_k <- f12_split %>% group_by(interval) %>% summarise(D = sum(event)) %>% select(D) %>% pull()

intervals <- unique(f12_split$interval) # number of intervals
fit_pois <- glm(D_k ~ offset(log(E_k))-1 + intervals, family = "poisson")
summary(fit_pois)
```


Hazards are the transformed coefficients, 

```{r}
exp(coef(fit_pois))
```

and you can get the standard errors from the output, too. To get the approximate SEs around the hazards rates, use the delta method:

```{r}
sqrt(diag(vcov(fit_pois)))*exp(coef(fit_pois))
```
## Question 4

Confirm that the estimated hazards from the regression are the same as the rates of birth in each interval implied by the data.

Answer: We could confirm this is the case.
```{r}
print(D_k/E_k)

print(exp(coef(fit_pois)))
```


## Visualizing hazards

In the lecture, I made a step-wise plot to visualize these hazards. The first step to get this is to make a tibble with our hazard rates, SEs and cut points. I add an extra point at the end, representing the maximum time observed:

```{r}
C <- length(cutpoints)+1
cuts <- c(0,cutpoints,max(f12$next.ivl))
hazs <- c(exp(coef(fit_pois)), exp(coef(fit_pois))[C])
ses <- c(sqrt(diag(vcov(fit_pois)))*exp(coef(fit_pois)), sqrt(diag(vcov(fit_pois)))[C]*exp(coef(fit_pois))[C])

haz_df <- tibble(cut = cuts, haz = hazs, se = ses)
```

Next we want to make some 95% CIs and calculate the mid-point and end-point of each interval, for plotting purposes. 

```{r}
haz_df <- haz_df %>% 
  mutate(lower = haz - 2*se,
         upper = haz + 2*se,
         midpoints = cut  + (lead(cut, default = 0) - cut)/2, 
         endpoints = lead(cut, default = max(cut)))
```

Now plot!

```{r}
haz_long <- haz_df %>% 
  pivot_longer(-(haz:upper), values_to = "time", names_to = "point")

haz_long %>% 
  ggplot(aes(time, haz) ) + geom_line() +
  geom_ribbon(aes(ymin = lower, ymax = upper), alpha = 0.2) + 
  geom_point(aes(time, haz), data = haz_long %>% filter(point == "midpoints")) +
  geom_vline(xintercept = cutpoints, col = 2, alpha = 0.2, lty = 2) + 
  theme_bw(base_size = 14) + 
  ylab("hazard rate") + 
  ggtitle("Estimated hazard rate of second birth\nby years since first birth")
```


## Survival probabilities

Would be good to also transform these hazards into survival probabilities. Here's the start of a function that does this:

```{r}
survival_prob <- function(lambdas,
                     cuts, # start and end times that lambdas refers to, starting at 0 and ending at max
                     ## observation time of interest,
                     ## thus length is one more than length of lambda
                     neval = 100 # at how many points do you want to evaluate S(t) within each interval?
                     ){
  lengthintervals <- rep((cuts[-1] - cuts[-length(cuts)])/neval, each = neval)
  t_seq <- c(0, cumsum(lengthintervals))
  cumulative_hazard <- cumsum(lengthintervals*rep(lambdas, each  = neval))
  surv_probs <- c(1, exp(-cumulative_hazard)) # add a 1 at the start because everyone survives at the start
  return(tibble(time = t_seq, surv = surv_probs ))
}
```

## Question 5

Fill in the gaps above in the `survival_prob` function. (Note you will need to remove `eval = FALSE` again). 

Now use this to plot the survival function:

```{r}
lambdas <- exp(coef(fit_pois))
cuts <- c(0, cutpoints, max(f12_split$next.ivl))
df_surv <- survival_prob(lambdas = exp(coef(fit_pois)), 
              cuts = cuts)

ggplot(aes(time, surv), data = df_surv) + geom_line()+
  ggtitle("Proportion of women who \nhave not had their second birth by time (years)") + 
  xlab("years") + ylab("S(t)")+
  theme_bw(base_size = 14)
```
## Question 5b (BONUS, not required)
Using simulation based on estimated log hazards and SEs, calculate and plot the 95% CI around the survival curve above.

# PCH with covariates 

## Question 6

Rerun the PCH regression above but with `age_group` as a covariate (Note: probability easiest just to run the individual-level regression rather than the regression on the sums). 

```{r}
age_regression <- glm(event ~ offset(log(interval_length)) - 1 + interval + age_group, data = f12_split, family = poisson)

print(summary(age_regression))
```


## Question 7

Use the `survival_prob` function defined above to help you find the proportion of women aged less than 30 who have had their second birth within 5 years of their first birth.

## Answer:
From the table we could find about 94.6% of women aged less than 30 had their second birth within 5 years of their first birth.
```{r}
exp_coef <- exp(coef(age_regression))
df_surv <- survival_prob(lambdas = exp_coef[-1], cuts = cuts)

print(1 - 0.05409992)
```



